Analysing the both the human and property cost of storm events in the United States since 1996 
========================================================

## Synopsis

no more than 10 sentcences describing work

## Data procressing

The NOAA dataset was downloaded from here. The storm data was read directly from the .bz2 file format without unzipping. To work with dates in the data set the storms start date (BGN_DATE) was used after it had been converted.

```{r,echo=TRUE, cache=TRUE}
library(stringr)

storm.data <- read.csv("../stormdata.bz2", stringsAsFactors=FALSE)

storm.data$BGN_DATE <- as.Date(sub(" 0:00:00","",storm.data$BGN_DATE),format="%m/%d/%Y")
storm.data$EVTYPE <- toupper(storm.data$EVTYPE)
storm.data$EVTYPE <- str_trim(storm.data$EVTYPE)

```

In 1996 the NOAA started to capture more storm event types and standardise the data entries. with that in mind all data prior to 1996 was removed.

```{r,echo=TRUE, cache=TRUE}
post.1995 <- storm.data[storm.data$BGN_DATE >= as.Date("01/01/1996",format="%m/%d/%Y"),]

multiplier <- c(3,6,9)
names(multiplier) <- c("K","M","B")

post.1995$fixed.crop.damage <-  ifelse(post.1995$CROPDMG>0,post.1995$CROPDMG*10^as.numeric(multiplier[post.1995$CROPDMGEXP]),post.1995$CROPDMG)
post.1995$fixed.prop.damage <-  ifelse(post.1995$PROPDMG>0,post.1995$PROPDMG*10^as.numeric(multiplier[post.1995$PROPDMGEXP]),post.1995$PROPDMG)
```

```{r, echo=TRUE}
post.1995$EVTYPE <-  sub("WILD/FOREST FIRE","WILDFIRE",post.1995$EVTYPE)
post.1995$EVTYPE <-  sub("RIP CURRENTS","RIP CURRENT",post.1995$EVTYPE)
post.1995$EVTYPE <-  sub("HURRICANE/TYPHOON","HURRICANE",post.1995$EVTYPE)
post.1995$EVTYPE <-  sub("TYPHOON","HURRICANE",post.1995$EVTYPE)
post.1995$EVTYPE <-  sub("^TSTM","THUNDERSTORM",post.1995$EVTYPE)
post.1995$EVTYPE <-  sub("STORM SURGE/TIDE","STORM SURGE",post.1995$EVTYPE)
post.1995$EVTYPE <-  sub("EROSION/CSTL FLOOD","COSTAL EROSION",post.1995$EVTYPE)
```

```{r,echo=TRUE}
library(plyr)
storm.harmful <- ddply(post.1995,c("EVTYPE"),summarise,total.injuries=sum(INJURIES),total.fatalities=sum(FATALITIES),avg.injuries=mean(INJURIES),avg.fatalities=mean(FATALITIES),total.events=length(EVTYPE))
storm.harmful$total.harm <- storm.harmful$total.injuries + storm.harmful$total.fatalities
storm.harmful$avg.harm <- storm.harmful$total.harm/storm.harmful$total.events

storm.harmful <- storm.harmful[storm.harmful$total.harm > 0, ]
storm.harmful.t10 <- head(storm.harmful[order(storm.harmful$total.harm,decreasing=TRUE),],10)

```

```{r,echo=TRUE}
storm.cost <- ddply(post.1995,c("EVTYPE"),summarise,total.crop=sum(fixed.crop.damage),total.prop=sum(fixed.prop.damage),avg.crop=mean(fixed.crop.damage),avg.prop=mean(fixed.prop.damage),total.events=length(EVTYPE))
storm.cost$total.cost <- storm.cost$total.crop + storm.cost$total.prop
storm.cost$avg.total.cost <- storm.cost$total.cost/storm.cost$total.events
storm.cost <- storm.cost[storm.cost$total.cost > 0,]
storm.cost.t10 <- head(storm.cost[order(storm.cost$total.cost,decreasing=TRUE),],10)
```
## Results
```{r,echo=TRUE,fig.width=10}
library(ggplot2)
g <- ggplot(storm.harmful.t10, aes(avg.injuries, avg.fatalities, size=total.events, label=EVTYPE))
g <- g + geom_point(colour="#56B4E9", alpha=I(0.4)) + geom_text(size=3, hjust=0.5, vjust=1) + scale_y_log10() + scale_x_log10()
g + scale_size(range=c(5,50)) + theme(legend.position="none")
```

```{r,echo=TRUE,fig.width=10}
g <- ggplot(storm.cost.t10, aes(avg.prop, avg.crop, size=total.events, label=EVTYPE))
g <- g + geom_point(colour="#56B4E9", alpha=I(0.4)) + geom_text(size=3, hjust=0.3, vjust=1) + scale_y_log10() + scale_x_log10()
g + scale_size(range=c(5,50)) + theme(legend.position="none")
```
